A Formalization of Typed Lua

### 2015-06-14 doc/papers/paper/paper.tex
DLS paper: note that not all rules are shown (RC2)
DLS: formatting (RC3)
DLS paper: fixing typos

\begin{abstract}
Programmers often migrate from a dynamically typed
to a statically typed language when their simple scripts
evolve into complex programs. Optional type systems are
one way of having both static and dynamic typing
in the same language, while keeping its dynamically typed
semantics. This makes evolving a program from dynamic to
static typing a matter of describing the implied types that
it it using and adding annotations to make those types explicit.
Designing an optional type system for
an existing dynamically typed language is challenging, as its types should feel natural to programmers that are already familiar with
this language.

@@

Typed Lua~\cite{maidl2014tl} is an optional type system for
Lua that is rich enough to preserve some of the idioms
that Lua programmers are already familiar with, while
helping programmers structure large Lua programs and
catch errors at compile-time.

@@

Rule {\sc S-TableUtoC} for subtyping between {\em unique} and {\em closed} table types is more permissive, as the typing rules of Section~\ref{sec:tables} guarantee that a {\em unique}
table type can only appear on the left side of a subtyping
judgment if the reference to the table with the {\em unique}
tag is being destroyed. This rule allows with subtyping, depth subtyping on any field, not only immutable fields, as well as the addition of new fields and the merging of existing fields, as long as the types of the values of new and merged fields include $\Nil$.

The rationale for the permissiveness of rule {\sc S-TableUtoC} derives from the fact that reading a non-existing key just yields $\Nil$ instead of being an error in Lua semantics, plus the fact
that the type system guarantees that a table with a
{\em unique} type has no keys with a value that does not
belong to one of its key types, and does not have any
aliases. With these guarantees, a table with type $\{1{:}\String, 2{:}\String,
``x"{:}\Integer, ``y"{:}\Value\}_{unique}$ can be safely recast as a table with type 
$\{\Number{:}\String \cup \Nil, ``x"{:}\Number, ``z"{:}
\Number \cup \Nil\}_{closed}$ as long as it becomes
inaccessible with the original type, which Section~\ref{sec:tables} will show that is the case
for any type judgment where this subtyping rule is used.

@@

The type $\Any$ is trivially consistent with all types, and all types
are consistent with $\Any$. A type is also consistent with
itself. Two union types $F_1 \cup F_2$ and $F_3 \cup F_4$ are
consistent if either $F_1 \sim F_3$ and $F_2 \sim F_4$ or
$F_1 \sim F_4$ and $F_2 \sim F_3$. Two function types
$S_1 \rightarrow S_2$ and $S_3 \rightarrow S_4$ are consistent
if and only if $S_1 \rightarrow S_3$ and $S_2 \rightarrow S_4$. Two table types are consistent if and only if they have the same key types, and the value types of each key type are consistent. Two recursive types are consistent if and only if their contents are consistent assuming that their type variables are consistent. Unions of tuple types follow the same rules as unions of first-level types. Variadic types are consistent if and only if their base types are consistent. 
Finally, pair types $F_1 \times P_1$ and $F_2 \times P_2$ are
consistent if and only if $F_1 \sim F_2$ and $P_1 \sim P_2$.

@@

We present the typing rules as a deduction system for two typing relations, one for typing statements and another for typing expressions. Typing judgments for statements have the form $\env_{1}, \penv \vdash s, \env_{2}$, meaning that
given a type environment $\env_{1}$ and a {\em projection environment} $\penv$ we can prove that a statement $s$ produces a new type environment $\env_{2}$. Typing judgments for
expressions have the form $\env_{1}, \penv \vdash e : T, \env_{2}$, meaning that given a type environment $\env_{1}$
and a projection environment $\penv$ we can prove that an expression $e$ has type $T$ and produces a new type environment $\env_{2}$.

@@

\begin{figure*}
	\[
	\begin{array}{rcl}
	wf(\{\overline{K:V}\}_{unique|open|fixed|closed}) & = & \forall i \; ((\nexists j \; i \not= j \,\wedge\, K_{i} \lesssim K_{j}) \,\wedge\, wf(V_{i}) \,\wedge\\
	& & \;\;\;\; \lnot tag(V_{i},unique) \,\wedge\, \lnot tag(V_{i},open))\\
	wf({\bf const}\; F) & = & wf(F) \\
	wf(F_1 \cup F_2) & = & wf(F_1) \,\wedge\, wf(F_2) \\
	wf(\mu x.F) & = & wf(F) \\
	wf(S_1 \rightarrow S_2) & = & wf(S_1) \,\wedge\, wf(S_2) \\
	wf(S_1 \sqcup S_2) & = & wf(S_1) \,\wedge\, S_2\\
	wf(F{*}) &=& wf(F) \\
	wf(F \times P) &=& wf(F) \,\wedge\,wf(P)\\
	wf(F) & = & \top \;\;\;\mathrm{for\; all\; other\; cases}\\\\
	tag(F_1 \cup F_2, t) & = & tag(F_1, t) \,\vee\, tag(F_2,t) \\
	tag(\{\overline{K:V}\}_t, t) & = & \top\\
	\multicolumn{3}{c}{
		\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{other\;recusive\;cases\;are\;similar,\;other\;base\;cases\;are\;}\bot}\\
\\\\
	fix(\{\overline{K:V}\}_{unique|open}) & = &
	\{\overline{K:V}\}_{fixed}\\
	\multicolumn{3}{c}{\;\;\;\;\;\;\;\mathrm{other\;cases\;by\;straightforward\;structural\;recursion}}\\\\
	merge(\overline{\env}) & = & reduce(\overline{\env}, merge2)\\
	merge2(\env_1,\env_2) & = & \{\overline{(id,merget(\env_1(id),\env_2(id))}\}\\
	merget(T_1, T_2) & = & T_1 \;\;\; \mathrm{if\;} T_2 \subtype T_1\\
	merget(T_1, T_2) & = & T_2 \;\;\; \mathrm{if\;} T_1 \subtype T_2\\
	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{the\;next\;case\;applies\;if}\;\forall i\;V^l_i\subtype_u V^r_i \,\vee\,V^r_i\subtype_u V^l_i \;\mathrm{and\;the\;right\; side\; is\;}wf}\\
	merget(\{\overline{K:V^l},\overline{K^{\prime}:V^\prime}\}_{unique},
	\{\overline{K:V^r},\overline{K^{\prime\prime}:V^{\prime\prime}}\}_{unique}) &=&
	\{\overline{K:sup_u(V^l,V^r)},\overline{K^\prime:V^\prime},\overline{K^{\prime\prime}:V^{\prime\prime}}\}_{unique}
	\\
	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{the\;next\;case\;applies\;if}\;\forall i\;V^l_i\subtype_c V^r_i \,\vee\,V^r_i\subtype_c V^l_i
		\;\mathrm{and\;the\;right\; side\; is\;}wf}\\
	merget(\{\overline{K:V^l},\overline{K^{\prime}:V^{\prime}}\}_{unique|open},
	\{\overline{K:V^r},\overline{K^{\prime\prime}:V^{\prime\prime}}\}_{unique|open}) &=&
	\{\overline{K:sup_c(V^l,V^r)},\overline{K^\prime:V^\prime},\overline{K^{\prime\prime}:V^{\prime\prime}}\}_{open}
	\\
	merget(T_1, T_2) & = & \bot \;\;\; \mathrm{otherwise}\\\\
	join(\env_1,\env_2) & = & \{\overline{(id,joint(\env_1(id),\env_2(id))}\}\\
	joint(T_1, T_2) & = & T_1 \;\;\; \mathrm{if\;} T_2 \subtype T_1\\
	joint(T_1, T_2) & = & T_2 \;\;\; \mathrm{if\;} T_1 \subtype T_2\\
	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{the\;next\;case\;applies\;if}\;\forall i\;V^l_i\subtype_u V^r_i \,\vee\,V^r_i\subtype_u V^l_i
		\;\mathrm{and\;the\;right\; side\; is\;}wf}\\
	joint(\{\overline{K:V^l},\overline{K^{\prime}:V^{\prime}}\}_{unique},
	\{\overline{K:V^r},\overline{K^{\prime\prime}:V^{\prime\prime}}\}_{unique}) &=&
	\{\overline{K:sup_u(V^l,V^r)},\overline{K^\prime:V^\prime\cup\Nil},\overline{K^{\prime\prime}:V^{\prime\prime}\cup\Nil}\}_{unique}\\
	joint(T_1, T_2) & = & \bot \;\;\; \mathrm{otherwise}\\\\
	sup_u(V_1, V_2) & = & V_2 \;\;\; \mathrm{if}\; V_1 \subtype_u V_2\\
	sup_u(V_1, V_2) & = & V_1 \;\;\; \mathrm{if}\; V_2 \subtype_u V_1\\\\
	reopen(\{\overline{K:V}\}_{closed}) & = & \{\overline{K:V}\}_{open} \\
	reopen(F) & = & F
	\end{array}
	\]
	\caption{Auxiliary Functions}
\label{fig:funcs}
\end{figure*}

@@

Open tables have a rule {\sc T-NewFieldOpen} that is not shown
in Figure~\ref{fig:tablerules} due to space. The only thing that
changes from one rule to the other is the tag in the type of $id$.

The type system needs to track aliasing of table references,
as the flexibility of {\em unique} and {\em open} tables
depend on respectively having no aliases and having only {\em closed} aliases. The system conservatively assumes that
a table has been aliased it its identifier appears in r-value position, with a couple of exceptions which we will explain shortly.

@@

Consider the following example, where we make an alias
of a {\em unique} table:
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{closed} = a \; \mathbf{in}a[``x"] \; {<}\String{>} = ``foo";$}
\end{tabular}
\end{center}

@@

The type $a$ has by the time of the inner {\bf local} block
is $\{``x":\String, ``y":\String\}_{unique}$. This type is
a subtype of the $\{``x":\String, ``y":\String\cup\Nil\}_{closed}$, as {\em unique} tables
have depth subtyping even in non-{\bf const} fields.
The coercion expression and the initialization of $b$
typecheck, and the type of $a$ changes to
$\{``x":\String, ``y":\String\cup\Nil\}_{open}$.
As the type is {\em open}, we can continue adding fields
to it, but are now restricted on future coercions. For
example, we will not be able to widen the type of $``x"$
to $\String \cup \Nil$, as that could break type safety
(by writing $\Nil$ to ``x" using reference $a$, and reading
``x" using the alias $b$).

@@

Closed-over variables that are not assigned still may
change their type if they have a {\em unique} table type
and are aliased inside the function body. We use the auxiliary
function $frv$ ({\em free referenced variables}) to collect
this set, and then function $openset$ changes any variables
with a {\em unique} table type in the environment to {\em open}.

Rule {\sc T-Function} types non-variadic function declarations,
and shows how these auxiliary functions work together.
Notice that the environment $\env_{2}$ resulting from
typing the function body is just ignored. The rule for
declaring variadic functions and recursive functions are
similar. Functions $closeall$, $closeset$, $openset$, $fav$,
and $frv$ are straightforward, and were omitted due to lack of space.

@@

The assignment $b=a$ typechecks because, at that point, 
both $b$ and $a$ have type $\{\}_{closed}$ due to $closeall$.
The set $fav$ is $\{b\}$, and the set $frv$ is $\{a\}$.
This causes the type of $b$ to also be $\{\}_{closed}$
in the body of the inner {\bf local} statement, due to $closeset$, and the type of $a$ to be $\{\}_{open}$ due
to $openset$. Being $\{\}_{open}$, adding a new field
is allowed, changing the type of $a$ to $\{``x": \Integer\}_{open}$ after the semicolon. After the call to
$f$, the variable $b$ is an alias to $a$ but this is not
an issue, as $\{\}_{closed}$ is a supertype of $\{``x": \Integer\}_{open}$.

@@

A standard way of decomposing the union in the assignment
would let the type of {\tt q} be $\Integer \times \Nil$ and
the type of {\tt r} be $\Integer \times \String$. Flow typing
would narrow the type of {\tt q} to $\Integer$ in the {\tt then}
branch and to $\Nil$ in the {\tt else} branch, but would
leave the type of {\tt r} unchanged.

{\em Projection types}
are a general way to make the dependency between the types
of {\tt q} and {\tt r} survive the decomposition of the tuple,
so flow typing can narrow all of the components by testing
just one of them.

@@

Several dynamically typed languages now have optional
type systems: Clojure~\cite{bonnaire-sergeant2012typed-clojure},
JavaScript~\cite{typescript}, Python~\cite{mypy}, 
and PHP~\cite{hack}. While Lua has some similarities to
all of these languages, none of these optional type
systems have the features described in this paper.
Most of these languages do not have idioms
that inspired these features; JavaScript has the
idiom of adding fields to an initially empty
record through assignment, but TypeScript sidesteps
the issue by having allowing an empty record to
have any record or object type, as the type of
missing fields is the bottom type of its type system.

@@

Our evaluation results show that our type system can statically
type check several Lua idioms and features, though the evaluation
also exposed several limitations.
We found that the three main limitations are
the lack of full support for overloaded functions, the lack
of parametric polymorphism, and operator overloading.
Overcoming these limitations is one major target for future work,
as it will allow us to statically type check more programs.

@@

Typed Lua has a working implementation\footnote{Available
at \url{http://github.com/andremm/typelua}.} that
Lua programmers can already use as a framework to
document, test, and structure their applications.
Even applications where a full conversion to static
type checking is unfeasible in the current state
of the type system, or too much work, can use Typed
Lua to document the external interfaces of its
libraries, giving the benefits of static typing
checking to the users of those libraries.

>>>>

\begin{abstract}
Programmers often migrate from a dynamically typed
to a statically typed language when their simple scripts
evolve into complex programs. Optional type systems are
one way of having both static and dynamic typing
in the same language, while keeping its dynamically typed
semantics. This makes evolving a program from dynamic to
static typing a matter of describing the implied types that
it is using and adding annotations to make those types explicit.
Designing an optional type system for
an existing dynamically typed language is challenging, as its types should feel natural to programmers that are already familiar with
this language.

@@

Typed Lua~\cite{maidl2014tl,maidl2015tl} is an optional type system for
Lua that is rich enough to preserve some of the idioms
that Lua programmers are already familiar with, while
helping programmers structure large Lua programs and
catch errors at compile-time.

@@

Rule {\sc S-TableUtoC} for subtyping between {\em unique} and {\em closed} table types is more permissive, as the typing rules of Section~\ref{sec:tables} guarantee that a {\em unique}
table type can only appear on the left side of a subtyping
judgment if the reference to the table with the {\em unique}
tag is being destroyed. This rule allows width subtyping, depth subtyping on any field, not only immutable fields, as well as the addition of new fields and the merging of existing fields, as long as the types of the values of new and merged fields include $\Nil$.

The rationale for the permissiveness of rule {\sc S-TableUtoC} derives from the fact that reading a non-existing key just yields $\Nil$ instead of being an error in Lua semantics, plus the fact
that the type system guarantees that a table with an
{\em unique} type has no keys with a value that does not
belong to one of its key types, and does not have any
aliases. With these guarantees, a table with type $\{1{:}\String, 2{:}\String,
``x"{:}\Integer, ``y"{:}\Value\}_{unique}$ can be safely recast as a table with type 
$\{\Number{:}\String \cup \Nil, ``x"{:}\Number, ``z"{:}
\Number \cup \Nil\}_{closed}$ as long as it becomes
inaccessible with the original type, which Section~\ref{sec:tables} will show that is the case
for any type judgment where this subtyping rule is used.

@@

The type $\Any$ is trivially consistent with all types, and all types
are consistent with $\Any$. A type is also consistent with
itself. Two union types $F_1 \cup F_2$ and $F_3 \cup F_4$ are
consistent if either $F_1 \sim F_3$ and $F_2 \sim F_4$ or
$F_1 \sim F_4$ and $F_2 \sim F_3$. Two function types
$S_1 \rightarrow S_2$ and $S_3 \rightarrow S_4$ are consistent
if and only if $S_1 \sim S_3$ and $S_2 \sim S_4$. Two table types are consistent if and only if they have the same key types, and the value types of each key type are consistent. Two recursive types are consistent if and only if their contents are consistent assuming that their type variables are consistent. Unions of tuple types follow the same rules as unions of first-level types. Variadic types are consistent if and only if their base types are consistent. 
Finally, pair types $F_1 \times P_1$ and $F_2 \times P_2$ are
consistent if and only if $F_1 \sim F_2$ and $P_1 \sim P_2$.

@@

We present the typing rules as a deduction system for two typing relations, one for typing statements and another for typing expressions. Typing judgments for statements have the form $\env_{1}, \penv \vdash s, \env_{2}$, meaning that
given a type environment $\env_{1}$ and a {\em projection environment} $\penv$ we can prove that a statement $s$ produces a new type environment $\env_{2}$. Typing judgments for
expressions have the form $\env_{1}, \penv \vdash e : T, \env_{2}$, meaning that given a type environment $\env_{1}$
and a projection environment $\penv$ we can prove that an expression $e$ has type $T$ and produces a new type environment $\env_{2}$.
Only the most relevant rules for the features that we are focusing on are shown.

@@

\begin{figure*}
	{\small
	\[
	\begin{array}{rcl}
	wf(\{\overline{K:V}\}_{unique|open|fixed|closed}) & = & \forall i \; ((\nexists j \; i \not= j \,\wedge\, K_{i} \lesssim K_{j}) \,\wedge\, wf(V_{i}) \,\wedge\\
	& & \;\;\;\; \lnot tag(V_{i},unique) \,\wedge\, \lnot tag(V_{i},open))\\
	wf({\bf const}\; F) & = & wf(F) \\
	wf(F_1 \cup F_2) & = & wf(F_1) \,\wedge\, wf(F_2) \\
	wf(\mu x.F) & = & wf(F) \\
	wf(S_1 \rightarrow S_2) & = & wf(S_1) \,\wedge\, wf(S_2) \\
	wf(S_1 \sqcup S_2) & = & wf(S_1) \,\wedge\, wf(S_2)\\
	wf(F{*}) &=& wf(F) \\
	wf(F \times P) &=& wf(F) \,\wedge\,wf(P)\\
	wf(F) & = & \top \;\;\;\mathrm{for\; all\; other\; cases}\\\\
	tag(F_1 \cup F_2, t) & = & tag(F_1, t) \,\vee\, tag(F_2,t) \\
	tag(\{\overline{K:V}\}_t, t) & = & \top\\
	\multicolumn{3}{c}{
		\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{other\;recusive\;cases\;are\;similar,\;other\;base\;cases\;are\;}\bot}\\
\\\\
	fix(\{\overline{K:V}\}_{unique|open}) & = &
	\{\overline{K:V}\}_{fixed}\\
	\multicolumn{3}{c}{\;\;\;\;\;\;\;\mathrm{other\;cases\;by\;straightforward\;structural\;recursion}}\\\\
	merge(\overline{\env}) & = & reduce(\overline{\env}, merge2)\\
	merge2(\env_1,\env_2) & = & \{\overline{(id,merget(\env_1(id),\env_2(id))}\}\\
	merget(T_1, T_2) & = & T_1 \;\;\; \mathrm{if\;} T_2 \subtype T_1\\
	merget(T_1, T_2) & = & T_2 \;\;\; \mathrm{if\;} T_1 \subtype T_2\\
	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{the\;next\;case\;applies\;if}\;\forall i\;V^l_i\subtype_u V^r_i \,\vee\,V^r_i\subtype_u V^l_i \;\mathrm{and\;the\;right\; side\; is\;}wf}\\
	merget(\{\overline{K:V^l},\overline{K^{\prime}:V^\prime}\}_{unique},
	\{\overline{K:V^r},\overline{K^{\prime\prime}:V^{\prime\prime}}\}_{unique}) &=&
	\{\overline{K:sup_u(V^l,V^r)},\overline{K^\prime:V^\prime},\overline{K^{\prime\prime}:V^{\prime\prime}}\}_{unique}
	\\
	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{the\;next\;case\;applies\;if}\;\forall i\;V^l_i\subtype_c V^r_i \,\vee\,V^r_i\subtype_c V^l_i
		\;\mathrm{and\;the\;right\; side\; is\;}wf}\\
	merget(\{\overline{K:V^l},\overline{K^{\prime}:V^{\prime}}\}_{unique|open},
	\{\overline{K:V^r},\overline{K^{\prime\prime}:V^{\prime\prime}}\}_{unique|open}) &=&
	\{\overline{K:sup_c(V^l,V^r)},\overline{K^\prime:V^\prime},\overline{K^{\prime\prime}:V^{\prime\prime}}\}_{open}
	\\
	merget(T_1, T_2) & = & \bot \;\;\; \mathrm{otherwise}\\\\
	join(\env_1,\env_2) & = & \{\overline{(id,joint(\env_1(id),\env_2(id))}\}\\
	joint(T_1, T_2) & = & T_1 \;\;\; \mathrm{if\;} T_2 \subtype T_1\\
	joint(T_1, T_2) & = & T_2 \;\;\; \mathrm{if\;} T_1 \subtype T_2\\
	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{the\;next\;case\;applies\;if}\;\forall i\;V^l_i\subtype_u V^r_i \,\vee\,V^r_i\subtype_u V^l_i
		\;\mathrm{and\;the\;right\; side\; is\;}wf}\\
	joint(\{\overline{K:V^l},\overline{K^{\prime}:V^{\prime}}\}_{unique},
	\{\overline{K:V^r},\overline{K^{\prime\prime}:V^{\prime\prime}}\}_{unique}) &=&
	\{\overline{K:sup_u(V^l,V^r)},\overline{K^\prime:V^\prime\cup\Nil},\overline{K^{\prime\prime}:V^{\prime\prime}\cup\Nil}\}_{unique}\\
	joint(T_1, T_2) & = & \bot \;\;\; \mathrm{otherwise}\\\\
	sup_u(V_1, V_2) & = & V_2 \;\;\; \mathrm{if}\; V_1 \subtype_u V_2\\
	sup_u(V_1, V_2) & = & V_1 \;\;\; \mathrm{if}\; V_2 \subtype_u V_1\\\\
	reopen(\{\overline{K:V}\}_{closed}) & = & \{\overline{K:V}\}_{open} \\
	reopen(F) & = & F
	\end{array}
	\]}
	\caption{Auxiliary Functions}
\label{fig:funcs}
\end{figure*}

@@

Open tables have a rule {\sc T-NewFieldOpen} that is not shown
in Figure~\ref{fig:tablerules} due to brevity. The only thing that
changes from one rule to the other is the tag in the type of $id$.

The type system needs to track aliasing of table references,
as the flexibility of {\em unique} and {\em open} tables
depend on respectively having no aliases and having only {\em closed} aliases. The system conservatively assumes that
a table has been aliased if its identifier appears in r-value position, with a couple of exceptions which we will explain shortly.

@@

Consider the following example, where we make an alias
of a {\em unique} table:
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{closed} = a \; \mathbf{in}$}\\
& & $\;a[``x"] \; {<}\String{>} = ``foo";$
\end{tabular}
\end{center}

@@

The type $a$ has by the time of the inner {\bf local} block
is $\{``x":\String, ``y":\String\}_{unique}$. This type is
a subtype of the $\{``x":\String, ``y":\String\cup\Nil\}_{closed}$, as {\em unique} tables
have depth subtyping even in non-{\bf const} fields.
The coercion expression and the initialization of $b$
typecheck, and the type of $a$ changes to
$\{``x":\String, ``y":\String\cup\Nil\}_{open}$.
As the type is {\em open}, we can continue adding fields
to it, but are now restricted on future coercions. For
example, we will not be able to widen the type of $``x"$
to $\String \cup \Nil$, as that could break type safety
(by writing $\Nil$ to $``x"$ using reference $a$, and reading
$``x"$ using the alias $b$).

@@

Closed-over variables that are not assigned still may
change their type if they have an {\em unique} table type
and are aliased inside the function body. We use the auxiliary
function $frv$ ({\em free referenced variables}) to collect
this set, and then function $openset$ changes any variables
with an {\em unique} table type in the environment to {\em open}.

Rule {\sc T-Function} types non-variadic function declarations,
and shows how these auxiliary functions work together.
Notice that the environment $\env_{2}$ resulting from
typing the function body is just ignored. The rule for
declaring variadic functions and recursive functions are
similar. Functions $closeall$, $closeset$, $openset$, $fav$,
and $frv$ are straightforward, and were omitted for brevity.

@@

The assignment $b=a$ typechecks because, at that point, 
both $b$ and $a$ have type $\{\}_{closed}$ due to $closeall$.
The set $fav$ is $\{b\}$, and the set $frv$ is $\{a\}$.
This causes the type of $b$ to also be $\{\}_{closed}$
in the body of the inner {\bf local} statement, due to $closeset$, and the type of $a$ to be $\{\}_{open}$ due
to $openset$. Being $\{\}_{open}$, adding a new field
is allowed, changing the type of $a$ to $\{``x": \Integer\}_{open}$ after the last semicolon. After the call to
$f$, the variable $b$ is an alias to $a$ but this is not
an issue, as $\{\}_{closed}$ is a supertype of $\{``x": \Integer\}_{open}$.

@@

A standard way of decomposing the union in the assignment
would let the type of $q$be $\Integer \times \Nil$ and
the type of $r$ be $\Integer \times \String$. Flow typing
would narrow the type of $q$ to $\Integer$ in the {\bf then}
branch and to $\Nil$ in the {\bf else} branch, but would
leave the type of $r$ unchanged.

{\em Projection types}
are a general way to make the dependency between the types
of $q$ and $r$ survive the decomposition of the tuple,
so flow typing can narrow all of the components by testing
just one of them.

@@

Several dynamically typed languages now have optional
type systems: Clojure~\cite{bonnaire-sergeant2015typed-clojure},
JavaScript~\cite{typescript}, Python~\cite{mypy}, 
and PHP~\cite{hack}. While Lua has some similarities to
all of these languages, none of these optional type
systems have the features described in this paper.
Most of these languages do not have idioms
that inspired these features; JavaScript has the
idiom of adding fields to an initially empty
record through assignment, but TypeScript sidesteps
the issue by having allowing an empty record to
have any record or object type, as the type of
missing fields is the bottom type of its type system.

@@

Our evaluation results show that our type system can statically
type check several Lua idioms and features, though the evaluation
also exposed several limitations.
We found that the three main limitations are the lack of
full support for overloaded functions, parametric polymorphism,
and operator overloading.
Overcoming these limitations is one major target for future work,
as it will allow us to statically type check more programs.

@@

Typed Lua has a working implementation\footnote{Available at
\url{https://github.com/andremm/typedlua}.}
that Lua programmers can already use as a framework to
document, test, and structure their applications.
Even applications where a full conversion to static
type checking is unfeasible in the current state
of the type system, or too much work,
Lua programmers can still use Typed Lua to document the
external interfaces of their libraries, giving the benefits of static typing
checking to the users of those libraries.
